%%
%% Commands for TeXCount
%TC:macro \cite [option:text,text]
%TC:macro \citep [option:text,text]
%TC:macro \citet [option:text,text]
%TC:envir table 0 1
%TC:envir table* 0 1
%TC:envir tabular [ignore] word
%TC:envir displaymath 0 word
%TC:envir math 0 word
%TC:envir comment 0 0
%%
%% For submission and review of your manuscript please change the
%% command to \documentclass[manuscript, screen, review]{acmart}.
%%
%% When submitting camera ready or to TAPS, please change the command
%% to \documentclass[sigconf]{acmart}.
%%
\documentclass[sigconf,authordraft, anonymous]{acmart}

\AtBeginDocument{%
  \providecommand\BibTeX{{Bib\TeX}}}

%% Rights management information.
%% Update these with your actual rights form values.
\setcopyright{acmlicensed}
\copyrightyear{2026}
\acmYear{2026}
\acmDOI{XXXXXXX.XXXXXXX}
\acmConference[ICCAD '26]{International Conference on Computer-Aided Design}{November 8--12, 2026}{San Jose, California, USA}
\acmISBN{978-1-4503-XXXX-X/2025/00}

%% Remove the copyright statement and ACM reference block from the PDF
\settopmatter{printacmref=false}
\renewcommand\footnotetextcopyrightpermission[1]{}
\pagestyle{plain}

\raggedbottom
\usepackage{lipsum}
\usepackage{wasysym} 
\usepackage{booktabs}
\usepackage{makecell}
\usepackage{svg}
\usepackage{listings}
\usepackage{xcolor}

\definecolor{codebg}{HTML}{F5F5F5}
\definecolor{codeframe}{HTML}{CCCCCC}
\definecolor{keyword}{HTML}{0060A8}
\definecolor{string}{HTML}{A31515}
\definecolor{comment}{HTML}{6A737D}

\lstdefinestyle{taidl}{
  language=Python,
  basicstyle=\ttfamily\scriptsize,
  keywordstyle=\color{keyword}\bfseries,
  stringstyle=\color{string},
  commentstyle=\color{comment}\itshape,
  backgroundcolor=\color{codebg},
  frame=single,
  rulecolor=\color{codeframe},
  framesep=4pt,
  xleftmargin=6pt,
  xrightmargin=6pt,
  aboveskip=6pt,
  belowskip=6pt,
  breaklines=true,
  columns=fullflexible,
  keepspaces=true,
  showstringspaces=false,
  tabsize=2,
  morekeywords={add_data_model, add_instruction, add_semantics,
                convert, dot, reshape, transpose},
  literate=
    {<-}{{$\leftarrow$}}2
    {->}{{$\rightarrow$}}2
    {\%A}{{{\color{keyword}\%A}}}2
    {\%B}{{{\color{keyword}\%B}}}2
    {\%C}{{{\color{keyword}\%C}}}2,
  captionpos=b,
}


%%
%% Author information
%%
\begin{document}

\title{TensorLift: Automatic Extraction of Tensor-Level ISA Semantics from Accelerator RTL via MLIR Semantic Lifting}

\author{Ruijie Gao}
\email{ruijieg@umich.edu}
\affiliation{%
  \institution{University of Michigan}
  \city{Ann Arbor}
  \state{Michigan}
  \country{United State of America}
}

\renewcommand{\shortauthors}{Gao et al.}

\begin{abstract}

Numerous tensor accelerator designs have been proposed in academia and industry, yet most lack well-documented ISAs and compiler backends. As a consequence, the majority of new designs are only evaluated on a handful of operators or synthetic kernels. Recent work (e.g., TAIDL and the ACT ecosystem) has shown that given a tensor-level ISA specification, complete software stacks including test oracles and compiler backends can be automatically generated. However, writing such specifications remains a manual, expert-driven process.

We present TensorLift, a fully automated framework that extracts tensor-level ISA specifications directly from accelerator RTL. Building on prior architecture-level model extraction that produces bit-level LLVM IR, TensorLift introduces an 8-pass MLIR semantic lifting pipeline that progressively recovers high-level tensor structure, including MAC idioms, saturation semantics, multi-dimensional buffer organizations, and data layout transformations. The pipeline emits specifications in the TAIDL formalism, immediately enabling automatic software stack generation through the ACT ecosystem.

We evaluate TensorLift on Gemmini, a systolic-array accelerator, extracting all 11 hardware instructions across 127 MLIR files. The lifting pipeline achieves up to 92.9\% reduction on processing elements---where bit-manipulation canonicalization collapses hundreds of sign-extension operations---and 30--40\% on controller modules whose remaining lines encode irreducible control logic (FSMs, address computation, muxing). TensorLift discovers hardware features omitted from the hand-written reference, including pooling engine semantics and im2col hardware support. Correctness is validated through SMT equivalence proofs and simulator testing. TensorLift closes the loop from RTL to full software stack without manual ISA documentation.
\end{abstract}

% \begin{CCSXML}
% <ccs2012>
%  <concept>
%   <concept_id>00000000.0000000.0000000</concept_id>
%   <concept_desc>Replace with correct CCS terms</concept_desc>
%   <concept_significance>500</concept_significance>
%  </concept>
% </ccs2012>
% \end{CCSXML}

% \ccsdesc[500]{Replace with correct CCS terms}

\keywords{Tensor Accelerators, ISA Extraction, RTL Abstraction, MLIR, Semantic Lifting, Hardware-Software Interface}

\maketitle


\input{sections/introduction}
\input{sections/background}
\input{sections/pipeline}
\input{sections/evaluation}
\input{sections/related}
\input{sections/conclusion}

\begin{acks}
% TODO: Add acknowledgments
\end{acks}

\clearpage

\bibliographystyle{ACM-Reference-Format}
\bibliography{references}

\end{document}
