\section{Background}
\label{sec:background}

% TODO: Write background


\subsection{Hardware-Software Interfaces in Tensor Accelerators}
\label{subsec:hw-sw-interface}

Tensor accelerators occupy a fundamentally different point in the design space than scalar or vector processors.
Rather than executing fine-grained instruction streams over flat register files and cache-coherent memory, they expose \emph{coarse-grained} operations that consume and produce multi-dimensional tensors and replace implicit caches with \emph{explicitly managed} on-chip memories (scratchpads, accumulator banks).
Gemmini~\cite{gemmini-dac2021}, Berkeley's open-source systolic-array generator, is representative: it attaches to a Rocket core~\cite{asanovic2016rocket} via the Rocket Custom Coprocessor (RoCC) interface~\cite{chipyard} with non-standard RISC-V custom instructions, and its programmer-visible state consists of a row-addressed scratchpad and an accumulator.
Its ISA exposes only configuration, DMA data-movement, and matrix-multiply execution instructions---every operation simultaneously specifies \emph{what} to compute, \emph{where} operands live in a private memory hierarchy, and \emph{how} data should be staged.

This distinctiveness places heavy demands on compiler infrastructure.
ML frameworks feed tensor IR (StableHLO, Linalg, TOSA) into backends such as XLA~\cite{xla}, TVM~\cite{tvm}, or IREE~\cite{iree}, which must lower high-level operations to the accelerator's native instructions.
Doing so correctly requires tiling and fusing loop nests to fit scratchpad capacity, scheduling explicit DMA transfers, allocating scratchpad regions, and orchestrating double-buffering---all of which presuppose a precise model of what each instruction does to which piece of tensor state.

Consider lowering a single XLA \texttt{dot} $C{=}A{\cdot}B$ for matrices exceeding Gemmini's $16{\times}16$ tile.
The backend must emit a tiled sequence: \texttt{config\_ex}/\texttt{config\_ld}/\texttt{config\_st} to set dataflow and stride parameters; \texttt{mvin}/\texttt{mvin2} to stage tiles of $A$ and $B$ into the scratchpad; \texttt{preload} followed by \texttt{compute\_preloaded} (or \texttt{compute\_accumulated}) to fire the systolic array for each output tile; and \texttt{mvout} to drain results.
Generating this sequence requires the backend to know, for every instruction, which tensor tile it touches, which scratchpad region it occupies, and how its output composes with subsequent operations.

This is precisely where most accelerators fall short.
Mapping tensor IR to an instruction sequence presupposes a \emph{machine-readable} ISA that specifies both the binary encoding and the \emph{tensor-level semantics} of each instruction.
For Gemmini, semantics live informally in prose documentation, the Chisel RTL, and a hand-written Spike functional model; most open-source accelerators offer only RTL and perhaps a driver.
Building a backend today therefore remains largely an exercise in \emph{manual reverse-engineering of the RTL}.

\subsection{Automated Compiler Generation and the Specification Bottleneck}
\label{subsec:spec-bottleneck}

Recent work has shown that the main barrier to end-to-end software support for tensor accelerators is not only backend engineering effort, but the lack of a precise, machine-readable Instruction Set Architecture (ISA) specification. TAIDL~\cite{taidl-micro2025} (Tensor Accelerator ISA Definition Language) addresses this problem by providing a domain-specific language for describing accelerator data models and instruction semantics at the tensor level. A key design choice in TAIDL is that instruction behavior is expressed using tensor operators close to XLA High Level Optimizer (XLA-HLO)~\cite{xla}, such as \texttt{reshape}, \texttt{transpose}, \texttt{convert}, \texttt{dot}, and \texttt{add}. This abstraction level is high enough to capture programmer-visible tensor semantics compactly, while remaining close enough to tensor compiler intermediate representations to support downstream automation.

A representative TAIDL specification therefore looks less like microarchitectural pseudocode and more like a tensor transformation. For example, a simplified multiply-accumulate style instruction can be expressed as follows:

\begin{figure}[h]
\begin{lstlisting}[style=taidl, caption={Simplified TAIDL specification for a Gemmini multiply-accumulate instruction (adapted from~\cite{taidl-ae-micro2025}). Only the core tensor semantics are shown; the full specification additionally includes dataflow routing and state constraint logic.}, label={lst:taidl-example}]
# Data model: scratchpad and accumulator
acc.add_data_model("sp", "1024*DIM", f"{DIM}xs8")
acc.add_data_model("bias", f"{DIM}", f"{DIM}xs32")

# Instruction with register operands
instr = acc.add_instruction("matmul8_compute",
            ["rs1", "rs2"])
instr.add_semantics(f"""
  %A.8:  {DIM}x{DIM}xs8  <- sp[@a.rs1:..., 0:{DIM}];
  %B.8:  {DIM}x{DIM}xs8  <- sp[@a.rs2:..., 0:{DIM}];
  %D.8:  {DIM}x{DIM}xs8  <- sp[@s.rs1:..., 0:{DIM}];
  %A.32: {DIM}x{DIM}xs32 = convert(%A.8);
  %B.32: {DIM}x{DIM}xs32 = convert(%B.8);
  %D.32: {DIM}x{DIM}xs32 = convert(%D.8);
  %dot:  {DIM}x{DIM}xs32 = dot(%A.32, %B.32),
         lhs_contracting_dims={{1}},
         rhs_contracting_dims={{0}};
  %C.32: {DIM}x{DIM}xs32 = add(%dot, %D.32);
  %C.cl: {DIM}x{DIM}xs32 = clamp(-128, %C.32, 127);
  %C.8:  {DIM}x{DIM}xs8  = convert(%C.cl);
  %C.8 -> sp[@s.rs2, 0];
""")
\end{lstlisting}
\end{figure}

\noindent This example illustrates the semantic granularity expected by TAIDL: the author must specify the tensor buffers exposed by the accelerator, the instruction operands, and the exact tensor-level effect of the instruction, but not the underlying Register-Transfer Level (RTL) control or datapath implementation. In published TAIDL examples~\cite{taidl-ae-micro2025}, even more complex instructions such as Intel Advanced Matrix Extensions (Intel AMX)~\cite{amx} \texttt{tdpbusd} are modeled by composing layout transformations (\texttt{reshape}, \texttt{transpose}), type conversion, tensor contraction, accumulation, and write-back in the same style.

Built on top of such specifications, ACT~\cite{act-arxiv} (Accelerator Compiler Toolkit) automates what comes after the ISA description. Given a TAIDL specification, ACT automatically generates compiler backends by combining equality saturation for instruction selection with constraint-programming-based memory allocation. The result is an ISA-driven compilation flow in which much of the backend logic is synthesized rather than handwritten. This is particularly attractive for tensor accelerators, whose instructions often operate on parameterized tensor tiles and explicitly managed on-chip memories that are difficult to support using conventional backend-generation techniques.

However, the automation boundary is also the key limitation: TAIDL and ACT automate everything downstream of the ISA specification, but they do not automate the creation of that specification itself---someone must still write the TAIDL model by hand, a task that requires deep knowledge of the hardware/software interface. This front-end specification burden is the gap that TensorLift addresses.

\subsection{RTL Extraction and MLIR-Based Semantic Lifting}
\label{subsec:rtl-extraction}

A natural baseline for recovering ISA semantics from hardware is architecture-level model extraction from RTL.
The relevant starting point is the autoGenILA line of work~\cite{autogenilaiccad, autogeniladate}, which first identifies the architectural state variables (ASVs) that persist across instructions and then derives the corresponding state update functions for each instruction directly from RTL.
This removes the need to manually write an Instruction-Level Abstraction (ILA)~\cite{ila} model and yields an architecture-level interface that is consistent with the hardware implementation by construction.
In TensorLift's workflow, autoGenILA serves exactly this role: it is the extraction backend that converts accelerator RTL into per-instruction semantic functions, providing the raw material for later lifting.

The problem is that these extracted models are still too low-level for tensor compiler generation.
autoGenILA captures hardware behavior faithfully, but the output remains a flattened, bit-level scalar model in which high-level tensor intent is obscured by the structure of the datapath~\cite{autogenilaiccad, autogeniladate}.
As illustrated in Figure~\ref{fig:code}, a single Gemmini processing element may initially appear as hundreds of primitive operations---dominated by sign-extension chains, truncations, bitwise composition, and scalar arithmetic---before later passes recover the underlying \texttt{dot\,+\,add\,+\,clamp} semantics.
Such an output is appropriate for bit-accurate reasoning, but it is not yet suitable for TAIDL~\cite{taidl-micro2025}, which requires tensor-visible semantics: buffer reads and writes, structured layout transformations, multiply-accumulate patterns, reductions, and saturation behavior.
In other words, autoGenILA solves the \emph{extraction} problem, but not the \emph{semantic abstraction} problem.

This is where MLIR (Multi-Level Intermediate Representation) becomes the key enabling substrate.
The advantage of MLIR is not any hardware-specific frontend, but its support for multi-level, progressive abstraction within one compiler framework~\cite{mlir-cgo2021}.
In TensorLift's implementation, the extracted semantics are materialized as low-level MLIR in the \texttt{arith} and \texttt{memref} dialects, and are progressively raised through structured passes that leverage MLIR's dialect-aware rewriting infrastructure~\cite{mlir-cgo2021, mlir-pattern-rewriting}.

More importantly, MLIR provides the right semantic stepping stones between bit-level arithmetic and tensor-level intent: \texttt{arith} for scalar operations, \texttt{memref} for explicit memory accesses, \texttt{scf} for structured control flow, and \texttt{linalg} for tensor computations~\cite{mlir-pattern-rewriting, mlir-partial-lowering, mlir-linalg-dialect}.
Because MLIR supports rewrite-driven and partially legalized transformations, TensorLift can progressively raise the representation rather than jumping from RTL-level behavior to tensor semantics in one step (Section~\ref{sec:pipeline}).
